---
title: "Practical Machine Learning Course - Final Project Report " 
author: "Subhdas"
date: "Saturday, June 20, 2015"
output: html_document
---

This is the final report for the Coursera course Practical Machine Learning under Data science specilization, from John Hopkins University. The code has been developed on a windows 64bit machine on windows 8.1 OS

**Introduction**

"Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement ??? a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset)."


**Data Sources**

The training data for this project are here:
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are here:
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

The data for this project comes from this original source: http://groupware.les.inf.puc-rio.br/har. If you use the document you create for this class for any purpose please cite them as they have been very generous in allowing their data to be used for this kind of assignment.

**Results Expected**

The goal of your project is to predict the manner in which they did the exercise. This is the "classe" variable in the training set. You may use any of the other variables to predict with. You should create a report describing how you built your model, how you used cross validation, what you think the expected out of sample error is, and why you made the choices you did. You will also use your prediction model to predict 20 different test cases.

Your submission should consist of a link to a Github repo with your R markdown and compiled HTML file describing your analysis. Please constrain the text of the writeup to < 2000 words and the number of figures to be less than 5. It will make it easier for the graders if you submit a repo with a gh-pages branch so the HTML page can be viewed online (and you always want to make it easy on graders :-).
You should also apply your machine learning algorithm to the 20 test cases available in the test data above. Please submit your predictions in appropriate format to the programming assignment for automated grading. See the programming assignment for additional details.

**Note from the Author**

Please refer to the r file, rmd file and the html document in file in the gitHub repo so as to understand the strategy adopted to predicted the class variable. Adequate measures has been taken to make sure Data is cleaned off of NA values before running the predictive analytics.

**Reproduceablity**

To get the same result as I obtained, you are supposed to install certain packages as mentioned in the code files. Most important is the caret package, which you can install by running the command, install.packages("caret").

The following Libraries were used for this project, which you must install - and load on your working environment.

```{r,echo=FALSE}
library(caret)
## Loading required package: lattice
## Loading required package: ggplot2
library(rpart)
library(rpart.plot)
library(RColorBrewer)
library(rattle)

## Rattle: A free graphical interface for data mining with R.
## Version 3.1.0 Copyright (c) 2006-2014 Togaware Pty Ltd.
## Type 'rattle()' to shake, rattle, and roll your data.
library(randomForest)
## randomForest 4.6-10


```

To set the same seed, run the following line of code.
```{r, echo=FALSE}
set.seed(12345)
trainUrl <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
testUrl <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"


```

**Getting the data**

The training data set and testing data set can be retrieved from the following URL:
```{r, echo=FALSE}
trainURL <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
testURL <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
```

Load the data from the URL's
```{r, echo=FALSE}
trainData <- read.csv(url(trainURL), na.strings=c("NA","#DIV/0!",""))
testData <- read.csv(url(testURL), na.strings=c("NA","#DIV/0!",""))
```

Use the following function to download the data
```{r, echo=FALSE}
#getDataFiles <- function(filesDirectory = "./") { 
#        if (!file.exists(filesDirectory)) {
#      dir.create(filesDirectory)
#           }
#        testUrl <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
#        trainUrl <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
#        trainFile <- "train.csv"
#        testFile <- "test.csv"
#   trainFilePath <- paste(filesDirectory, trainFile, sep = "/")
#        testFilePath <- paste(filesDirectory, testFile, sep = "/")
#        download.file(trainUrl, destfile = trainFilePath, method="curl")
#        download.file(testUrl, destfile = testFilePath, method="curl")
#        training <- read.csv(trainFilePath, na.strings=c("NA","#DIV/0!",""))
#        testing <- read.csv(testFilePath, na.strings=c("NA","#DIV/0!",""))
#}

#getDataFiles("/data")
```

**Make two partitions of the train Data Set**

Partioning Training data set into two data sets, 60% for myTraining, 40% for myTesting:
```{r, echo=FALSE}
inTrain <- createDataPartition(y=trainData$classe, p=0.6, list=FALSE)
myTraining <- trainData[inTrain, ]; myTesting <- trainData[-inTrain, ]
dim(myTraining); dim(myTesting)
```

**Cleaning the data**

The following transformations were used to clean the data:

Transformation 1: Cleaning NearZeroVariance Variables. Then run the following code to view possible NZV Variables:

```{r, echo=FALSE}
myDataNZV <- nearZeroVar(myTraining, saveMetrics=TRUE)

#Run this code to create another subset without NZV variables:

myNZVvars <- names(myTraining) %in% c("new_window", "kurtosis_roll_belt", "kurtosis_picth_belt",
"kurtosis_yaw_belt", "skewness_roll_belt", "skewness_roll_belt.1", "skewness_yaw_belt",
"max_yaw_belt", "min_yaw_belt", "amplitude_yaw_belt", "avg_roll_arm", "stddev_roll_arm",
"var_roll_arm", "avg_pitch_arm", "stddev_pitch_arm", "var_pitch_arm", "avg_yaw_arm",
"stddev_yaw_arm", "var_yaw_arm", "kurtosis_roll_arm", "kurtosis_picth_arm",
"kurtosis_yaw_arm", "skewness_roll_arm", "skewness_pitch_arm", "skewness_yaw_arm",
"max_roll_arm", "min_roll_arm", "min_pitch_arm", "amplitude_roll_arm", "amplitude_pitch_arm",
"kurtosis_roll_dumbbell", "kurtosis_picth_dumbbell", "kurtosis_yaw_dumbbell", "skewness_roll_dumbbell",
"skewness_pitch_dumbbell", "skewness_yaw_dumbbell", "max_yaw_dumbbell", "min_yaw_dumbbell",
"amplitude_yaw_dumbbell", "kurtosis_roll_forearm", "kurtosis_picth_forearm", "kurtosis_yaw_forearm",
"skewness_roll_forearm", "skewness_pitch_forearm", "skewness_yaw_forearm", "max_roll_forearm",
"max_yaw_forearm", "min_roll_forearm", "min_yaw_forearm", "amplitude_roll_forearm",
"amplitude_yaw_forearm", "avg_roll_forearm", "stddev_roll_forearm", "var_roll_forearm",
"avg_pitch_forearm", "stddev_pitch_forearm", "var_pitch_forearm", "avg_yaw_forearm",
"stddev_yaw_forearm", "var_yaw_forearm")
myTraining <- myTraining[!myNZVvars]
#To check the new N?? of observations
dim(myTraining)

```


Transformation 2: Killing first column of Dataset - ID Removing first ID variable so that it does not interfer with ML Algorithms:
```{r, echo=FALSE}
myTraining <- myTraining[c(-1)]
```

Transformation 3: Cleaning Variables with too many NAs. For Variables that have more than a 60% threshold of NA's I'm going to leave them out:

```{r, echo=FALSE}
trainingV3 <- myTraining #creating another subset to iterate in loop
for(i in 1:length(myTraining)) { #for every column in the training dataset
        if( sum( is.na( myTraining[, i] ) ) /nrow(myTraining) >= .6 ) { #if n?? NAs > 60% of total observations
        for(j in 1:length(trainingV3)) {
            if( length( grep(names(myTraining[i]), names(trainingV3)[j]) ) ==1)  { #if the columns are the same:
                trainingV3 <- trainingV3[ , -j] #Remove that column
            }   
        } 
    }
}
#To check the new N?? of observations
dim(trainingV3)
```

```{r, echo=FALSE}
#Seting back to our set:
myTraining <- trainingV3
rm(trainingV3)
#dim(trainingV3)
```


Now let us do the exact same 3 transformations but for our myTesting and testing data sets.

```{r, echo=FALSE}

clean1 <- colnames(myTraining)
clean2 <- colnames(myTraining[, -58]) #already with classe column removed
myTesting <- myTesting[clean1]
testData <- testData[clean2]

#To check the new N?? of observations
dim(myTesting)
```



```{r, echo=FALSE}
dim(testData)

#Note: The last column - problem_id - which is not equal to training sets, was also "automagically" removed
#No need for this code:
#testing <- testing[-length(testing)]
```

In order to ensure proper functioning of Decision Trees and especially RandomForest Algorithm with the Test data set (data set provided), we need to coerce the data into the same type.
```{r, echo=FALSE}
for (i in 1:length(testData) ) {
        for(j in 1:length(myTraining)) {
        if( length( grep(names(myTraining[i]), names(testData)[j]) ) ==1)  {
            class(testData[j]) <- class(myTraining[i])
        }      
    }      
}
#And to make sure Coertion really worked, simple smart ass technique:
testData <- rbind(myTraining[2, -58] , testData) #note row 2 does not mean anything, this will be removed right.. now:
testData <- testData[-1,]
```

**Using ML algorithms for prediction: Decision Tree**

```{r, echo=FALSE}

modelFIT0 <- rpart(classe ~ ., data=myTraining, method="class")

```


Note: to view the decision tree with fancy run this command:
```{r, echo=FALSE}
fancyRpartPlot(modelFIT0)
```

Predicting:
```{r, echo=FALSE}
predictALG <- predict(modelFIT0, myTesting, type = "class")
```


(Moment of truth) Using confusion Matrix to test results:
```{r, echo=FALSE}
confusionMatrix(predictALG, myTesting$classe)
```


**Using ML algorithms for prediction: Random Forests**


```{r, echo=FALSE}
modelFIT <- randomForest(classe ~. , data=myTraining)
```

Predicting in-sample error:
```{r, echo=FALSE}

predictRF1 <- predict(modelFIT, myTesting, type = "class")
```

(Moment of truth) Using confusion Matrix to test results:

```{r, echo=FALSE}

confusionMatrix(predictRF1, myTesting$classe)
```


Random Forests yielded better Results, as expected!

**Generating Files to submit as answers for the Assignment:**

Finally, using the provided Test Set out-of-sample error.

For Random Forests we use the following formula, which yielded a much better prediction in in-sample:
```{r, echo=FALSE}

predictRF2 <- predict(modelFIT, testData, type = "class")
```


**Code to : Generate a File**

```{r, echo=TRUE}
writeTable = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}

writeTable(predictRF2)
```

